RationalGRL: A Framework for Argumentation and Goal Modeling

### 2017-10-23 discussion.tex
related research finished

There are several contributions to the literature that relate argumentation-based techniques with goal modeling. The contribution most closely related to ours is the work by Jureta \emph{et al.}~\cite{Jureta:RE2008}. Jureta \emph{et al.} propose ``Goal Argumentation Method (GAM)'' to guide argumentation and justification of modeling choices during the construction of goal models. One of the elements of GAM is the translation of formal argument models to goal models (similar to ours). In this sense, our RationalGRL framework can be seen as an instantiation and implementation of  part of the GAM. The main difference between our approach and GAM is that we integrate arguments and goal models using argument schemes, and that we develop these argument schemes by analyzing transcripts. GAM instead uses structured argumentation.

In previous work on the RationalGRL framework~\cite{vanzee-etal:renext2015,vanZee-etal:er2016}, we have explored one way of combining PRAS and GRL. In this work, which takes a similar approach to \cite{Jureta:RE2008}, argument diagrams are translated to GRL models (an automatic translation tool is discussed in~\cite{vanZee-etal:comma2016}). The argument diagrams are complex practical reasoning arguments, where the premises are the goals and the conclusion is some action we need to perform to realize those goals. Thus, we have essentially two complex diagrams, a practical reasoning argument diagram and a GRL goal diagram, and a mapping between them.

Connecting argumentation and goal modelling by providing a mapping between two types of diagram was, in our opinion, ultimately an unsatisfying solution given the problems and requirements described in Section \ref{sect:introduction}. One problem is that the argument diagram is at least as complex as the GRL diagram, so any stakeholder trying to understand the discussion has to parse two complex diagrams containing goals, alternative solutions, tasks, and so forth. Furthermore, in our previous work we did not provide the specific critical questions for goal models (see Section \ref{sect:gmas}). This meant that any counterargument had to be constructed from scratch, as no guidance was given as to possible ways to criticize a GRL model. So the previous iterations of the RationalGRL framework violated requirement 1: argument diagrams do not closely mirror the actual discussions of stakeholders in which ideas are proposed and challenged. Furthermore, not having specific argument schemes and critical questions for goal modelling makes it hard to develop a guiding methodology for the use of argumentation in goal modelling (requirement 4).

One idea is to capture requirements engineering and software design processes as explicit dialogues between parties~\cite{finkelstein1989multiparty}.

\todo{F}{F}{Review the following:

\begin{itemize}
\item Mirbel and Villata on Dung \& goal models
\item Software design discussions Black et al., Prakken and Wierenga, \cite{finkelstein1989multiparty}
  \end{itemize}
}
\subsection{Open issues}
\label{sect:goalmodeling:openissues}

@@

>>>>

There are several contributions to the literature that relate argumentation-based techniques with goal modeling. The first line of work is by Bagrheri and Ensan \cite{bagheri2011consolidating} and Mirbel and Villata \cite{MirbelVillata12} use abstract Dung argumentation frameworks (see Section~\ref{sect:formalframework}) to capture individual goals and the relations between them as modeled in a goal model, so that consistent (i.e., acceptable) subsets of goals can be computed using the relevant argumentation semantics. For example, if in the goal model there is a conflict between goals $G_1$ and $G_2$, there is an attack between the arguments representing these goals and hence $G_1$ and $G_2$ cannot be in the same extension. Or if there is a dependency link between $G_1$ and $G_2$, then $G_1$ should be in every extension $G_2$ is in and vice versa. AND and OR decomposition are also modeled thus, that is, if $G_3$ AND-decomposes into $G_1$ and $G_2$ then $G_1$ and $G_2$ should be in every extension $G_3$ is in, and if $G_3$ OR-decomposes into $G_1$ and $G_2$ then either $G_1$ or $G_2$ should be in every extension $G_3$ is in.

Modeling goal models as argumentation frameworks allows one to compute the possible consistent sets of goals and tasks given a goal model. In RationalGRL, we have opted not to provide such an argumentation-theoretic semantics to goal models. The reason for this is that GRL already has quite fine-grained satisfiability semantics \cite{Amyot:2010:EGM:1841349.1841356}, which take into account conflicts, decompositions and dependencies. 
RationalGRL focuses on what is not captured in the work discussed above \cite{bagheri2011consolidating,MirbelVillata12}, namely the arguments and beliefs underlying a goal model, and the way these arguments and beliefs interact with the elements of the goal model. If desired, however, it would be easy to provide Dung semantics similar to \cite{bagheri2011consolidating,MirbelVillata12} for goal models, as the elements of a goal model are already arguments in RationalGRL (Figure~\ref{fig:metamodel}).

The contribution most closely related to ours is the work by Jureta \emph{et al.}~\cite{Jureta:RE2008}. Jureta \emph{et al.} propose ``Goal Argumentation Method (GAM)'' to guide argumentation and justification of modeling choices during the construction of goal models. GAM is a high-level decision process, in which alternative solutions for an RE problem are evaluated and compared using argumentation. Jureta et al. give a simple made-up example of an argumentative discussion regarding a meeting scheduler, in which a goal model is being built by the stakeholders proposing tasks, goals, and alternative solutions for goals. They include clarification as an important step in their GAM process, and discuss various types of clarity problems (vagueness, ambiguity, overgenerality, synonymy) and basic techniques for dealing with them (e.g. thesaurus checks, labeling vague expressions). Jureta et al. then present the argumentation part of their framework, where reasons (justifications) for conclusions are given as formal structured arguments\footnote{Informally, a structured argument is similar to the PRAS argument in Section~\ref{sect:background:pras}, i.e., $a, b \vec{therefore} c$.}. Arguments and alternatives are then captured as structured arguments or argument diagrams with reasons for or against goals and tasks. Given these arguments the set of undefeated (i.e., acceptable) propositions can be computed to determine which alternative solution to a problem is acceptable. Thus, the arguments and beliefs underlying a goal model and possible alternative modelings are captured as formal arguments. Furthermore, a mapping from goal models to argument diagrams is given, so that it is possible to start arguing about an already existing goal model.

The GAM process is essentially a generic, high-level process for problem solving, not specifically tailored towards goal modeling. In this sense, the RationalGRL methodology in Figure~\ref{fig:rationalgrl-methodology} can be seen as a more specific version of GAM explicitly meant for goal modeling. The argument schemes and critical questions in RationalGRL provide more clear handles for goal modeling (cf. requirement 4 of our framework, Section~\ref{sect:introduction}).

In previous work on the RationalGRL framework~\cite{vanzee-etal:renext2015,vanZee-etal:er2016}, we effectively extended Jureta et al.'s work and translated argument diagrams to GRL models\footenote{In \cite{Jureta:RE2008} only a mapping from goal models to structured arguments is given, and the step from structured arguments or argument diagrams to goal models is never formally defined.} (an automatic translation tool is discussed in~\cite{vanZee-etal:comma2016}). Thus, we have essentially two complex diagrams, an argument diagram and a goal diagram, and a mapping between them. This was, in our opinion, ultimately an unsatisfying solution given the problems and requirements described in Section \ref{sect:introduction}. One problem is that the argument diagram is at least as complex as the GRL diagram, so any stakeholder trying to understand the discussion has to parse two complex diagrams containing goals, alternative solutions, tasks, and so forth. So the previous iterations of the RationalGRL framework violated requirement 1: argument diagrams do not closely mirror the actual discussions of stakeholders in which ideas are proposed and challenged.

\subsection{Open issues}
\label{sect:goalmodeling:openissues}

@@

One idea is to capture requirements engineering and software design processes as explicit dialogues between parties~\cite{finkelstein1989multiparty}. Software design discussions Black et al., Prakken and Wierenga.
